---
title: "BS_Behavioral"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Box Sync/BaskingShark_behaviors")
ID <- 3
data <- read.csv(paste0("Shark ", ID, "_CATS_behaviors.csv"))

head(data, 20)
class(data$Time)
summary(data)


#clean up CATS Cam data so we only get what we need
#data <- data[,c("Time", "Pressure...channel..1", "Sway..g.", "Surge..g.", "Heave..g.", "Light.intensity...channel..1", "Temperature.ext....channel..1", "ODBA")] 

# need to convert data values to meters
head(data)
data$Pressure...channel..1 <- (data$Pressure...channel..1*.001)+.3
summary(data$Pressure...channel..1) # got this information from Heather Vance


# This is at 20 Hz. We want to extract every 20th value and get rid of milliseconds
data2 <- data[seq(1, nrow(data), 20),]
vector <- as.character(data2$Time)
data2$Time <- substr(vector, 1, 8)
head(data2$Time)
head(data2)


# make into chron object so that we can merge the behavioral data according to timestamp
# needed to make as a character first so that chron recognizes the format
#install.packages("chron")
library(chron)
data2$chron<- chron(times. = data2$Time, format ="hh:mm:ss", out.format = "h:m:s")
data2$chron
class(data2$chron)



## need to keep magnetometer reading, but which one? 
# let's search out the signal 

# data.trash <- data2[200:400,]
# summary(data.trash)
# nrow(data.trash)
# summary(data.trash$Pressure...channel..1)
# plot(data.trash$Compass..1....channel..1~data.trash$chron, type="l")
# plot(data.trash$Compass..1....channel..2~data.trash$chron, type="l")
# plot(data.trash$Compass..1....channel..3~data.trash$chron, type="l")
# plot(data.trash$Gyroscope..2....channel..1~data.trash$chron, type="l")
# plot(data.trash$Gyroscope..2....channel..2~data.trash$chron, type="l")
# plot(data.trash$Gyroscope..2....channel..3~data.trash$chron, type="l")
# 
# plot(data2$Gyroscope..2....channel..3~data2$chron, type="l")
```



```{r}
##########################################
### let's read in the behavioral data now 
##########################################
behavior <- read.csv("CATS_S3_Events_csv.csv")
head(behavior)
class(behavior$Time)

# use this function (found on internet) to generate time values from the seconds values in the behavior data
dhms <- function(t){
    paste(t %/% (60*60*24) 
         ,paste(formatC(t %/% (60*60) %% 24, width = 2, format = "d", flag = "0")
               ,formatC(t %/% 60 %% 60, width = 2, format = "d", flag = "0")
               ,formatC(t %% 60, width = 2, format = "d", flag = "0")
               ,sep = ":"
               )
         )
}


behavior$Time <- dhms(behavior$Time)
head(behavior$Time)

# clean it up 
vector2 <- as.character(behavior$Time)
behavior$Time <- substr(vector2, 3, 10) #this allows you to get rid of objects within a character string (only want the 3-10 values in the character string)
head(behavior$Time)

# now make this into a chron object
behavec <- chron(times. = behavior$Time, format ="hh:mm:ss", out.format = "h:m:s")
head(behavec)

#### We want to add all of these times to the start time from the data
behavior$Time2 <- behavec + data2$chron[1]
head(behavior$Time2)
head(behavior)

## clean up data so that we only have behavior, comments and time2 columns
behavior2 <- behavior[,(3:5)]
head(behavior2)
colnames(behavior2) <- c("Behavior", "Comment", "chron")
head(behavior2)
```


```{r}
##############################
# Check out all of the columns in both datasets to make sure we can merge them # 
############################
# CATS data
head(data2) # going to merge using the "chron" column
# behavior data
head(behavior2) #going to merge using "chron" column

# now to join 
library(tidyverse)
library(dplyr)
fulldat <- left_join(data2, behavior2, by= c("chron"="chron"))
head(fulldat)
fulldat <- fulldat[,-1]
head(fulldat, 200)
summary(fulldat)
#note that accelerometer values are in g
colnames(fulldat) <- c("Time2", "Pressure", "Compass1", "Compass2", "Compass3", "AccelSway", "Sway.g.", "AccelSurge", "Surge.g.", "AccelHeave", "Heave.g.", "Temperature", "Gyro1", "Gyro2", "Gyro3", "Accel1", "Accel2", "Accel3", "Temp2", "Light", "TempExt", "ODBA", "Time", "Behavior", "Comments")
head(fulldat)

# get rid of the top rows before the behaviors begin
head(fulldat, 200)
fulldat <- fulldat[126:nrow(fulldat),]
head(fulldat)

# visualize the pressure data to make sure calibration makes sense
plot(fulldat$Time, (-fulldat$Pressure), type = "l")
# see an outlier at 86 m
fulldat <- fulldat[fulldat$Pressure<86,]
summary(fulldat$Pressure)
plot(fulldat$Time, (-fulldat$Pressure), type = "l")

head(fulldat)
##### Now have merged behaviors into the major data set 
```
Now that we have cleaned up our data and combined the CATS and behavioral data sets, we need to fill in the behaviors that are currently missing for every other timestamp (since we set it up so that we only had a behavior assigned to the time when it started)

```{r}


# for loop for filling in behaviors
for (i in 1:nrow(fulldat)) {
  if(is.na(fulldat[i,]$Behavior)==TRUE){
    fulldat[i,]$Behavior <- fulldat[(i-1),]$Behavior
  }
}

summary(fulldat)

##### Now to clean up data so that we only have depths below 2 meters
nrow(fulldat) # 12997

fulldat2 <- fulldat[abs(fulldat$Pressure)>=2,]
summary(fulldat2$Pressure)
plot(-(fulldat2$Pressure)~fulldat2$Time, type ="p", cex=.2)
nrow(fulldat2) #3541

summary(fulldat2)


######################## now divvy up data set so that we have the following ##############
#### if we keep it simple: feed vs. non-feeding

summary(fulldat2$Pressure)

# going to use fulldat2 to keep it below 2 m to get rid of surface movement
fulldat2$feeding <- ifelse (fulldat2$Behavior == "Feed"|fulldat2$Behavior == "Lturn while feed"| fulldat2$Behavior == "Rturn while feed"| fulldat2$Behavior == "Swallow", 1, 0)

summary(fulldat2$Pressure)
#so, feeding here = 1; non-feeding = 1

### surface swimming - i.e. swimming above 10 m but below 2 #####
summary(fulldat2$Behavior)
fulldat_ss <- fulldat2[abs(fulldat2$Pressure)<10 & fulldat2$Behavior!="Swim along ground",]
#chose 10 because that's the most shallow dive, but also excluded swimming along ground (shallower parts of the habitat may be less than or = to 10 m)
summary(fulldat_ss$Behavior)
fulldat_ss2<- fulldat_ss[fulldat_ss$feeding==0,]
nrow(fulldat_ss2)
plot(-(fulldat_ss2$Pressure)~fulldat_ss2$Time, type ="p", cex=.2)



### surface feeding - i.e. feeding above 3 m #####

fulldat_sf <- fulldat2[abs(fulldat2$Pressure)<10,]
fulldat_sf2<- fulldat_sf[fulldat_sf$feeding==1,]
summary(fulldat_sf2$Behavior)
nrow(fulldat_sf2)

plot(-(fulldat_sf2$Pressure)~fulldat_sf2$Time, type ="p", cex=.2)

### ground swimming - i.e. swimming below 26 m #####
fulldat_gs <- fulldat2[abs(fulldat2$Pressure)>=26 & fulldat2$Behavior=="Swim along ground",]
summary(fulldat_gs$Behavior)
fulldat_gs2<- fulldat_gs[fulldat_gs$feeding==0,]
nrow(fulldat_gs2)
plot(-(fulldat_gs2$Pressure)~fulldat_gs2$Time, type ="p", cex=.2)


```
** Data Cleaning Done** Time to divide the water column to look at relevant behaviors in each

```{r}



summary(fulldat)
write.csv(fulldat, file = "S3_everysec_behaviors.csv")
## Assuming each row is 1 second. Double check this:
(fulldat$Time[nrow(fulldat)] - fulldat$Time[1]) #[1] 03:36:22
nrow(fulldat)

### if each row is a second, then the number of times each behavior is assigned to a second should give us the duration of that behavior over time

```


Here we see that feeding time (vs. all other behaviors (swimming and turning)) decreases by depth, which makes intuitive sense. It would be interesting to consider the environmental context in which this occurs (likely feeding only in the plankton 
layer). 


**Now to do the same thing for Shark 2**

```{r}
setwd("~/Box Sync/BaskingShark_behaviors")
ID <- 2
data_2 <- read.csv(paste0("Shark ", ID, "_CATS_behaviors.csv"))

head(data_2, 20)
class(data_2$Time)
summary(data_2)

#clean up data so we only get what we need
data_2 <- data_2[,c("Time", "Pressure...channel..1", "Sway", "Surge", "Heave","Acceleration..2....channel..1","Acceleration..2....channel..2", "Acceleration..2....channel..3", "Light.intensity...channel..1", "Temperature.ext....channel..1", "ODBA")] 
  
 
# calculate pressure to meters #
data_2$Pressure...channel..1 <- (data_2$Pressure...channel..1*.001)+27.71
summary(data_2$Pressure...channel..1) #Heather did this pressure calculation

# This is at 20 Hz. We want to extract every 20th value and get rid of milliseconds
data2_s2 <- data_2[seq(1, nrow(data_2), 20),]
vector <- as.character(data2_s2$Time)
data2_s2$Time <- substr(vector, 1, 8)
head(data2_s2$Time)
head(data2_s2)
# make into chron object so that we can merge the behavioral data according to timestamp
# needed to make as a character first so that chron recognizes the format
#install.packages("chron")
library(chron)
data2_s2$chron<- chron(times. = data2_s2$Time, format ="hh:mm:ss", out.format = "h:m:s")
data2_s2$chron
class(data2_s2$chron)
```


```{r}
##########################################
### let's read in the behavioral data now 
##########################################
behavior_s2 <- read.csv("CATS_S2_Events.csv")
head(behavior_s2)
class(behavior_s2$Time)

# use this function (found on internet) to generate time values from the seconds values in the behavior data
dhms <- function(t){
    paste(t %/% (60*60*24) 
         ,paste(formatC(t %/% (60*60) %% 24, width = 2, format = "d", flag = "0")
               ,formatC(t %/% 60 %% 60, width = 2, format = "d", flag = "0")
               ,formatC(t %% 60, width = 2, format = "d", flag = "0")
               ,sep = ":"
               )
         )
}


behavior_s2$Time <- dhms(behavior_s2$Time)
head(behavior_s2$Time)

# clean it up 
vector2_s2 <- as.character(behavior_s2$Time)
behavior_s2$Time <- substr(vector2_s2, 3, 10) #this allows you to get rid of objects within a character string (only want the 3-10 values in the character string)
head(behavior_s2$Time)

# now make this into a chron object
behavec_s2 <- chron(times. = behavior_s2$Time, format ="hh:mm:ss", out.format = "h:m:s")
head(behavec_s2)

#### We want to add all of these times to the start time from the data
behavior_s2$Time2 <- behavec_s2 + data2_s2$chron[1]
head(behavior_s2$Time2)
head(behavior_s2)
class(behavior_s2$Time2)
## clean up data so that we only have behavior, comments and time2 columns
behavior2_s2 <- behavior_s2[,(2:4)]
head(behavior2_s2)
colnames(behavior2_s2) <- c("Behavior", "Comment", "chron")
head(behavior2_s2)

```

```{r}
##############################
# Check out all of the columns in both datasets to make sure we can merge them # 
############################
# CATS data
head(data2_s2) # going to merge using the "chron" column
# behavior data
head(behavior2_s2) #going to merge using "chron" column

# now to join 
library(tidyverse)
library(dplyr)
fulldat_s2 <- left_join(data2_s2, behavior2_s2, by= c("chron"="chron"))
head(fulldat_s2)
fulldat_s2 <- fulldat_s2[,-1]
head(fulldat_s2, 200)

summary(fulldat_s2)
#note that accelerometer values are in g
colnames(fulldat_s2) <- c("Pressure", "Sway", "Surge", "Heave", "Acc2.1", "Acc2.2", "Acc2.3", "Light", "Temperature", "ODBA", "chron", "Behavior", "comments")
head(fulldat_s2)

# get rid of the top rows before the behaviors begin
head(fulldat_s2, 500)
fulldat_s2 <- fulldat_s2[294:nrow(fulldat_s2),]
head(fulldat_s2)

## plot pressure to check that calibrations worked ##
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l")
summary(fulldat_s2$Pressure) ## see some outliers at 90 m 
class(fulldat_s2$Pressure)
fulldat_s2 <- fulldat_s2[(fulldat_s2$Pressure < 90), ]
summary(fulldat_s2$Pressure)
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l")
##### Now have merged behaviors into the major data set 
```
Now that we have cleaned up our data and combined the CATS and behavioral data sets, we need to fill in the behaviors that are currently missing for every other timestamp (since we set it up so that we only had a behavior assigned to the time when it started)

```{r}
# for loop for filling in behaviors
for (i in 1:nrow(fulldat_s2)) {
  if(is.na(fulldat_s2[i,]$Behavior)==TRUE){
    fulldat_s2[i,]$Behavior <- fulldat_s2[(i-1),]$Behavior
  }
}

summary(fulldat_s2)
head(fulldat_s2)

##### Now to clean up data so that we only have depths below 2 meters
nrow(fulldat_s2) # 10147

fulldat2.2 <- fulldat_s2[abs(fulldat_s2$Pressure)>=2,]
summary(fulldat2.2$Pressure)
plot(-(fulldat2.2$Pressure)~fulldat2.2$chron, type ="p", cex=.2)
nrow(fulldat2.2) #4304

summary(fulldat2)


######################## now divvy up data set so that we have the following ##############
#### if we keep it simple: feed vs. non-feeding

summary(fulldat2.2$Pressure)

# going to use fulldat2 to keep it below 2 m to get rid of surface movement
fulldat2.2$feeding <- ifelse (fulldat2.2$Behavior == "Feed"|fulldat2.2$Behavior == "Lturn while feed"| fulldat2.2$Behavior == "Rturn while feed"| fulldat2.2$Behavior == "Swallow", 1, 0)

summary(fulldat2.2$Pressure)
#so, feeding here = 1; non-feeding = 1

### surface swimming - i.e. swimming above 12 m but below 2 #####
summary(fulldat2.2$Behavior)
fulldat_ss.2 <- fulldat2.2[abs(fulldat2.2$Pressure)<10 & fulldat2.2$Behavior!="Swim along bottom",]
#chose 12 because that's the most shallow dive, but also excluded swimming along ground (shallower parts of the habitat may be less than or = to 12 m)
summary(fulldat_ss.2$Behavior)
fulldat_ss2.2<- fulldat_ss.2[fulldat_ss.2$feeding==0,]
nrow(fulldat_ss2.2)
plot(-(fulldat_ss2.2$Pressure)~fulldat_ss2.2$chron, type ="p", cex=.2)



### surface feeding - i.e. feeding above 3 m #####

fulldat_sf.2 <- fulldat2.2[abs(fulldat2.2$Pressure)<12,]
fulldat_sf2.2<- fulldat_sf.2[fulldat_sf.2$feeding==1,]
summary(fulldat_sf2.2$Behavior)
nrow(fulldat_sf2.2)
fulldat_sf2.2

plot(-(fulldat_sf2.2$Pressure)~fulldat_sf2.2$chron, type ="p", cex=.2)

### ground swimming - i.e. swimming below 26 m #####
summary(fulldat2.2)
summary(fulldat2.2$Pressure)
summary(fulldat2.2$Behavior)
fulldat_gs.2 <- fulldat2.2[abs(fulldat2.2$Pressure)>=16 & fulldat2.2$Behavior=="Swim along bottom",]
summary(fulldat_gs.2$Behavior)
fulldat_gs2.2<- fulldat_gs.2[fulldat_gs.2$feeding==0,]
nrow(fulldat_gs2.2) #1041
plot(-(fulldat_gs2.2$Pressure)~fulldat_gs2.2$chron, type ="p", cex=.2)

```
** Data Cleaning Done** Time to divide the water column to look at relevant behaviors in each

```{r}

#### if we keep it simple: feed vs. non-feeding
fulldat_s2$feeding <- ifelse (fulldat_s2$Behavior == "Feed"|fulldat_s2$Behavior == "Lturn while feed"| fulldat_s2$Behavior == "Rturn while feed"| fulldat_s2$Behavior == "Swallow", 1, 0)

summary(fulldat_s2$feeding)

summary(fulldat_s2)

write.csv(fulldat_s2, file = "S2_everysec_behaviors.csv")
# convert pressure (mB) to depth 
#fulldat$Pressure2 <- abs(fulldat$Pressure*0.010197442889221)
#range(fulldat$Pressure)

```

Looking at all behaviors as function of depth (**Figure 1**)

```{r}
ftrash <- fulldat[fulldat$feeding==1, ]
ftrash_s2 <- fulldat_s2[fulldat_s2$feeding==1, ] #includes swallowing
swi <- fulldat[fulldat$feeding==0, ]
swi_s2 <- fulldat_s2[fulldat_s2$feeding==0, ] #all non-feeding
cons <- fulldat[fulldat$Behavior=="Conspecific interaction",]
cons_s2 <- fulldat_s2[fulldat_s2$Behavior=="Conspecific interaction",]


plot(fulldat$Time, -(fulldat$Pressure), type="l", col="black", main="S3 Depth Profile + Feed/NF", xlab="Time", ylab="Depth (m)")
abline(v=c(ftrash$Time), col = rgb(red=0, green=0, blue = 1, alpha = 0.01)) # blue = feed
abline(v=c(swi$Time), col = rgb(red=1, green=0, blue = 0, alpha = 0.01)) # rd = non-feed
abline(v=c(cons$Time), col = rgb(red=0, green=1, blue = 0, alpha = 0.01)) #green = conspecific interaction

# white = areas where the camera was out of water
#alpha allows us to make these transparent

# now do it on shark 2
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l", col="black", main="S2 Depth Profile + Feed/NF", xlab="Time", ylab="Depth (m)")
abline(v=c(ftrash_s2$chron), col = rgb(red=0, green=0, blue = 1, alpha = 0.01)) # blue = feed
abline(v=c(swi_s2$chron), col = rgb(red=1, green=0, blue = 0, alpha = 0.01)) # rd = non-feed
abline(v=c(cons_s2$chron), col = rgb(red=0, green=1, blue = 0, alpha = 0.01)) #green = conspecific interaction

```
***Accelerometer data*
We will need to run a low pass filter over the tailbeat data (our "sway" axis)

```{r}
#install.packages("signal")
library(signal)

#s3
plot(fulldat$Heave~fulldat$Time, type = "l", main="Accelerometer signals, S3", xlab="Acceleration (m/s^3)", ylab="Time")
lines(fulldat$Surge~fulldat$Time, col="red")
lines(fulldat$Sway~fulldat$Time, col="blue")

#s2
plot(fulldat_s2$Sway~fulldat_s2$chron, type = "l", main="Accelerometer signals, S2", xlab="Acceleration (m/s^3)", ylab="Time")
lines(fulldat_s2$Surge~fulldat_s2$chron, col="red")
lines(fulldat_s2$Heave~fulldat_s2$chron, col="blue")


# to get rid of high frequency signals in the sway axis, need to run a low pass filter. Note that we already started to do that by only getting a data point every second (vs. 20/second)

#based on what we see on the S3 video, we have approximately one TB signal every 3 seconds
# 
# ## what if we subset every 3 seconds?
# fulldat_swayset <- fulldat[seq(1, nrow(fulldat), 3),]
# head(fulldat_swayset)
# 
# plot(fulldat_swayset$Sway[1:50]~fulldat_swayset$Time[1:50], type = "l")

```
Our key metrics are going to be TB frequency and speed
```{r}
# Practice bringing out speed, which is in the magnetometer component of the data because that's where the paddlewheel was attached
```
