---
title: "BS_Behavioral"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Box Sync/BaskingShark_behaviors")
ID <- 3
data <- read.csv(paste0("Shark ", ID, "_CATS_behaviors.csv"))

head(data, 20)
class(data$Time)
summary(data)

#clean up CATS Cam data so we only get what we need
data <- data[,c("Time", "Pressure...channel..1", "Sway..g.", "Surge..g.", "Heave..g.", "Light.intensity...channel..1", "Temperature.ext....channel..1", "ODBA")] 

# need to convert data values to meters
head(data)
data$Pressure...channel..1 <- (data$Pressure...channel..1*.001)+.3
summary(data$Pressure...channel..1) # got this information from Heather...need to confirm (86 meters?)


# This is at 20 Hz. We want to extract every 20th value and get rid of milliseconds
data2 <- data[seq(1, nrow(data), 20),]
vector <- as.character(data2$Time)
data2$Time <- substr(vector, 1, 8)
head(data2$Time)
head(data2)
### this gives us t

# make into chron object so that we can merge the behavioral data according to timestamp
# needed to make as a character first so that chron recognizes the format
#install.packages("chron")
library(chron)
data2$chron<- chron(times. = data2$Time, format ="hh:mm:ss", out.format = "h:m:s")
data2$chron
class(data2$chron)

```



```{r}
##########################################
### let's read in the behavioral data now 
##########################################
behavior <- read.csv("CATS_S3_Events_csv.csv")
head(behavior)
class(behavior$Time)

# use this function (found on internet) to generate time values from the seconds values in the behavior data
dhms <- function(t){
    paste(t %/% (60*60*24) 
         ,paste(formatC(t %/% (60*60) %% 24, width = 2, format = "d", flag = "0")
               ,formatC(t %/% 60 %% 60, width = 2, format = "d", flag = "0")
               ,formatC(t %% 60, width = 2, format = "d", flag = "0")
               ,sep = ":"
               )
         )
}


behavior$Time <- dhms(behavior$Time)
head(behavior$Time)

# clean it up 
vector2 <- as.character(behavior$Time)
behavior$Time <- substr(vector2, 3, 10) #this allows you to get rid of objects within a character string (only want the 3-10 values in the character string)
head(behavior$Time)

# now make this into a chron object
behavec <- chron(times. = behavior$Time, format ="hh:mm:ss", out.format = "h:m:s")
head(behavec)

#### We want to add all of these times to the start time from the data
behavior$Time2 <- behavec + data2$chron[1]
head(behavior$Time2)
head(behavior)

## clean up data so that we only have behavior, comments and time2 columns
behavior2 <- behavior[,(3:5)]
head(behavior2)
colnames(behavior2) <- c("Behavior", "Comment", "chron")
head(behavior2)
```


```{r}
##############################
# Check out all of the columns in both datasets to make sure we can merge them # 
############################
# CATS data
head(data2) # going to merge using the "chron" column
# behavior data
head(behavior2) #going to merge using "chron" column

# now to join 
library(tidyverse)
library(dplyr)
fulldat <- left_join(data2, behavior2, by= c("chron"="chron"))
head(fulldat)
fulldat <- fulldat[,-1]
head(fulldat, 200)

#note that accelerometer values are in g
colnames(fulldat) <- c("Pressure", "Sway", "Surge", "Heave", "Light", "Temperature", "ODBA", "Time", "Behavior", "Comments")
head(fulldat)

# get rid of the top rows before the behaviors begin
head(fulldat, 200)
fulldat <- fulldat[126:nrow(fulldat),]
head(fulldat)

# visualize the pressure data to make sure calibration makes sense
plot(fulldat$Time, (-fulldat$Pressure), type = "l")
# see an outlier at 86 m
fulldat <- fulldat[fulldat$Pressure<86,]
summary(fulldat$Pressure)
plot(fulldat$Time, (-fulldat$Pressure), type = "l")
##### Now have merged behaviors into the major data set 
```
Now that we have cleaned up our data and combined the CATS and behavioral data sets, we need to fill in the behaviors that are currently missing for every other timestamp (since we set it up so that we only had a behavior assigned to the time when it started)

```{r}


# for loop for filling in behaviors
for (i in 1:nrow(fulldat)) {
  if(is.na(fulldat[i,]$Behavior)==TRUE){
    fulldat[i,]$Behavior <- fulldat[(i-1),]$Behavior
  }
}

summary(fulldat)


### need to get rid of all behaviors that are "out of water"
fulldat <- fulldat[fulldat$Behavior!="Out of water",]
summary(fulldat$Behavior) #note: behavior is still kept here as a level for the factor, which should be fine 

### plot pressure values from new dataset without "out of water" behaviors

plot(fulldat$Time, (-fulldat$Pressure), type = "l")
### We now have a data set that contains all behaviors ###
```
** Data Cleaning Done** Time to divide the water column to look at relevant behaviors in each

```{r}

#### if we keep it simple: feed vs. non-feeding
fulldat$feeding <- ifelse (fulldat$Behavior == "Feed"|fulldat$Behavior == "Lturn while feed"| fulldat$Behavior == "Rturn while feed"| fulldat$Behavior == "Swallow", 1, 0)

#so, feeding here = 1; non-feeding = 1

## Assuming each row is 1 second. Double check this:
(fulldat$Time[nrow(fulldat)] - fulldat$Time[1]) #[1] 03:36:22
nrow(fulldat)

### if each row is a second, then the number of times each behavior is assigned to a second should give us the duration of that behavior over time

```


Here we see that feeding time (vs. all other behaviors (swimming and turning)) decreases by depth, which makes intuitive sense. It would be interesting to consider the environmental context in which this occurs (likely feeding only in the plankton 
layer). 


**Now to do the same thing for Shark 2**

```{r}
setwd("~/Box Sync/BaskingShark_behaviors")
ID <- 2
data_2 <- read.csv(paste0("Shark ", ID, "_CATS_behaviors.csv"))

head(data_2, 20)
class(data_2$Time)
summary(data_2)

#clean up data so we only get what we need
data_2 <- data_2[,c("Time", "Pressure...channel..1", "Sway", "Surge", "Heave","Acceleration..2....channel..1","Acceleration..2....channel..2", "Acceleration..2....channel..3", "Light.intensity...channel..1", "Temperature.ext....channel..1", "ODBA")] 
  
 
# calculate pressure to meters #
data_2$Pressure...channel..1 <- (data_2$Pressure...channel..1*.001)+27.71
summary(data_2$Pressure...channel..1) #Heather did this pressure calculation

# This is at 20 Hz. We want to extract every 20th value and get rid of milliseconds
data2_s2 <- data_2[seq(1, nrow(data_2), 20),]
vector <- as.character(data2_s2$Time)
data2_s2$Time <- substr(vector, 1, 8)
head(data2_s2$Time)
head(data2_s2)
# make into chron object so that we can merge the behavioral data according to timestamp
# needed to make as a character first so that chron recognizes the format
#install.packages("chron")
library(chron)
data2_s2$chron<- chron(times. = data2_s2$Time, format ="hh:mm:ss", out.format = "h:m:s")
data2_s2$chron
class(data2_s2$chron)
```


```{r}
##########################################
### let's read in the behavioral data now 
##########################################
behavior_s2 <- read.csv("CATS_S2_Events.csv")
head(behavior_s2)
class(behavior_s2$Time)

# use this function (found on internet) to generate time values from the seconds values in the behavior data
dhms <- function(t){
    paste(t %/% (60*60*24) 
         ,paste(formatC(t %/% (60*60) %% 24, width = 2, format = "d", flag = "0")
               ,formatC(t %/% 60 %% 60, width = 2, format = "d", flag = "0")
               ,formatC(t %% 60, width = 2, format = "d", flag = "0")
               ,sep = ":"
               )
         )
}


behavior_s2$Time <- dhms(behavior_s2$Time)
head(behavior_s2$Time)

# clean it up 
vector2_s2 <- as.character(behavior_s2$Time)
behavior_s2$Time <- substr(vector2_s2, 3, 10) #this allows you to get rid of objects within a character string (only want the 3-10 values in the character string)
head(behavior_s2$Time)

# now make this into a chron object
behavec_s2 <- chron(times. = behavior_s2$Time, format ="hh:mm:ss", out.format = "h:m:s")
head(behavec_s2)

#### We want to add all of these times to the start time from the data
behavior_s2$Time2 <- behavec_s2 + data2_s2$chron[1]
head(behavior_s2$Time2)
head(behavior_s2)
class(behavior_s2$Time2)
## clean up data so that we only have behavior, comments and time2 columns
behavior2_s2 <- behavior_s2[,(2:4)]
head(behavior2_s2)
colnames(behavior2_s2) <- c("Behavior", "Comment", "chron")
head(behavior2_s2)

```

```{r}
##############################
# Check out all of the columns in both datasets to make sure we can merge them # 
############################
# CATS data
head(data2_s2) # going to merge using the "chron" column
# behavior data
head(behavior2_s2) #going to merge using "chron" column

# now to join 
library(tidyverse)
library(dplyr)
fulldat_s2 <- left_join(data2_s2, behavior2_s2, by= c("chron"="chron"))
head(fulldat_s2)
fulldat_s2 <- fulldat_s2[,-1]
head(fulldat_s2, 200)

summary(fulldat_s2)
#note that accelerometer values are in g
colnames(fulldat_s2) <- c("Pressure", "Sway", "Surge", "Heave", "Acc2.1", "Acc2.2", "Acc2.3", "Light", "Temperature", "ODBA", "chron", "Behavior", "comments")
head(fulldat_s2)

# get rid of the top rows before the behaviors begin
head(fulldat_s2, 500)
fulldat_s2 <- fulldat_s2[294:nrow(fulldat_s2),]
head(fulldat_s2)

## plot pressure to check that calibrations worked ##
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l")
summary(fulldat_s2$Pressure) ## see some outliers at 90 m 
class(fulldat_s2$Pressure)
fulldat_s2 <- fulldat_s2[(fulldat_s2$Pressure < 90), ]
summary(fulldat_s2$Pressure)
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l")
##### Now have merged behaviors into the major data set 
```
Now that we have cleaned up our data and combined the CATS and behavioral data sets, we need to fill in the behaviors that are currently missing for every other timestamp (since we set it up so that we only had a behavior assigned to the time when it started)

```{r}
# for loop for filling in behaviors
for (i in 1:nrow(fulldat_s2)) {
  if(is.na(fulldat_s2[i,]$Behavior)==TRUE){
    fulldat_s2[i,]$Behavior <- fulldat_s2[(i-1),]$Behavior
  }
}

summary(fulldat_s2)
head(fulldat_s2)

### need to get rid of all behaviors that are "out of water"
fulldat_s2 <- fulldat_s2[fulldat_s2$Behavior!="Out of water",]
summary(fulldat_s2$Behavior) #note: behavior is still kept here as a level for the factor, which should be fine 

## now plot pressure without "out of water" behaviors
plot(fulldat_s2$chron, (-fulldat_s2$Pressure), type = "l")
### We now have a data set that contains all behaviors ###
```
** Data Cleaning Done** Time to divide the water column to look at relevant behaviors in each

```{r}

#### if we keep it simple: feed vs. non-feeding
fulldat_s2$feeding <- ifelse (fulldat_s2$Behavior == "Feed"|fulldat_s2$Behavior == "Lturn while feed"| fulldat_s2$Behavior == "Rturn while feed"| fulldat_s2$Behavior == "Swallow", 1, 0)

summary(fulldat_s2$feeding)
# convert pressure (mB) to depth 
#fulldat$Pressure2 <- abs(fulldat$Pressure*0.010197442889221)
#range(fulldat$Pressure)



```

Looking at all behaviors as function of depth

```{r}
ftrash <- fulldat[fulldat$feeding==1, ]
ftrash_s2 <- fulldat_s2[fulldat_s2$feeding==1, ] #includes swallowing
swi <- fulldat[fulldat$feeding==0, ]
swi_s2 <- fulldat_s2[fulldat_s2$feeding==0, ] #all non-feeding
cons <- fulldat[fulldat$Behavior=="Conspecific interaction",]
cons_s2 <- fulldat_s2[fulldat_s2$Behavior=="Conspecific interaction",]


plot(fulldat$Time, -(fulldat$Pressure), type="l", col="black", main="S3 Depth Profile + Feed/NF", xlab="Time", ylab="Depth (m)")
abline(v=c(ftrash$Time), col = rgb(red=0, green=0, blue = 1, alpha = 0.01)) # blue = feed
abline(v=c(swi$Time), col = rgb(red=1, green=0, blue = 0, alpha = 0.01)) # rd = non-feed
abline(v=c(cons$Time), col = rgb(red=0, green=1, blue = 0, alpha = 0.01)) #green = conspecific interaction

# white = areas where the camera was out of water
#alpha allows us to make these transparent

# now do it on shark 2
plot(fulldat_s2$chron, -(fulldat_s2$Pressure), type="l", col="black", main="S2 Depth Profile + Feed/NF", xlab="Time", ylab="Depth (m)")
abline(v=c(ftrash_s2$chron), col = rgb(red=0, green=0, blue = 1, alpha = 0.01)) # blue = feed
abline(v=c(swi_s2$chron), col = rgb(red=1, green=0, blue = 0, alpha = 0.01)) # rd = non-feed
abline(v=c(cons_s2$chron), col = rgb(red=0, green=1, blue = 0, alpha = 0.01)) #green = conspecific interaction

```

