---
title: "MatLab-R conversion_CATS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Shark_speed
%% work out speed of shark using magnetometer signal and
% rotations of the magnet fitted on the external paddle wheel.

```{r}
setwd("~/Box Sync/BaskingShark_behaviors")
ID <- 3
data <- read.csv(paste0("Shark ", ID, "_CATS_behaviors.csv"))
head(data, 100)
M.data <- data$Compass..1....channel..3
class(M.data)
M.samplingrate <- 20 #20 samples/second
t = (1:length(M.data))/M.samplingrate
plot(t[1:100], M.data[1:100], type="l")
```

% comp_filt seperates the signal into two frequency bands
% Here we are high-pass filtering the data at around half of the average
% cycles per second, which I worked out to be about 4 for the paddle wheel.

**Need to download signal package in R to do this**

```{r}
#X=comp_filt(M.data,M.sampling_rate,2);
#Mh=X{2};

library(signal)
fft.Mdata <- fft(M.data)
plot(Mod(fft.Mdata), type='l', main='Raw serie - fft spectrum')

### The following code removes the first spike in the spectrum
### This would be the high pass filter
inx_filter = 15
N = length(M.data)
FDfilter = rep(1, N)
FDfilter[1:inx_filter] = 0
FDfilter[(N-inx_filter):N] = 0
fft.y_noise_filtered = FDfilter * fft.Mdata
plot(Mod(fft.y_noise_filtered), type='l', main='HP Filtered')


### creating a Butterworth filter for data (used because maximally flat, order of 5 = steepest cutoff; order of 1 = shallowest cut-off)
# used W = .5 because 1/2 the average cycles per second (?)
# z for digital filter
bf<- butter(n=5, W=.5, type = "high", plane = "z")


# now applying the filter to the data
z <- filter(bf, M.data)

plot(z[1:100], type="l")
```
